# Generalized Linear Model

For the exercises in this section, we require following package, data and functions. You can download them from canvas.

```{r, echo = T, message=FALSE}
library(nnet)
load("_data/taxonomy.rdata")
load("_data/species.rdata")
source("_functions/CV.class.multinom.R")
```

## Species Data

Consider the plant species count data used in [lecture of Generalized linear model](https://nmbu.instructure.com/courses/483/files/220998?module_item_id=20284).

- a. Redo the analysis using the `Biomass` and `pH` variables as predictors for species counts (as in slide 9). Find separate expressions for the estimated linear predictors $\hat{\eta}_j$ for each `pH`-level $j$ = "`low`","`medium`","`high`". Also find the estimated expected count for `Biomass = 2` and `pH = high`.

<div class="ans">

First we fit the model:

```{r}
#options(contrasts=c("contr.treatment","contr.poly"))
mod2.glm <- glm(Species ~ Biomass + pH, family = poisson, data = species)
summod2 <- summary(mod2.glm)
summod2$coefficients
```

Then the estimated linear predictors are:

$$\eta_\text{low} = 3.85 - 0.128\cdot \text{Biomass} - 1.14$$

$$\eta_\text{medium} = 3.85 - 0.128\cdot \text{Biomass} - 0.45$$

$$\eta_\text{high} = 3.85 - 0.128\cdot \text{Biomass}$$

For `Biomass = 2` and `pH = high` the estimated expected count is:

$$\hat{\lambda} = \exp(3.85 - 0.128\cdot 2) = 36.4$$

Alternatively we could compute this in R by:

```{r}
predict(mod2.glm, newdata = data.frame(Biomass = 2, pH = "high"), type = "response")
```

</div>

- b. Fit the same model as in a. using the quasipoisson approach. Is there any evidence of overdisperion?

<div class="ans">

```{r}
mod2q.glm <- glm(Species ~ Biomass + pH, family = quasipoisson, data = species)
summary(mod2q.glm)
```

The dispersion parameter is only marginally larger than 1. We conclude that over-disperion is not a problem here.

</div>

- c. Extend the model from a. by including an interaction between Biomass and pH. Test the significance of the interaction. Is there a significant difference between pH="high" and pH="medium" with regard to the effect of Biomass to the species count?

<div class="ans">

```{r}
mod3.glm <- glm(Species ~ Biomass * pH, family = poisson, data = species)
summary(mod3.glm)
anova(mod2.glm, mod3.glm, test = "Chi")
```

The interaction term is highly significant with a p-value of 0.00033. The Wald tests (the approximate normal test) given in the summary output gave a p-value of 0.167 for the term `Biomass:pHmid` which really is the contrast between interactions `Biomass:pHhigh`and `Biomass:pHmedium` since `pHhigh` is the reference level in this model. The relatively large p-value says that there is no evidence that Biomass has different effect on counts for these two pH-levels.
</div>


- d. Optional R-programming exercise for those interested: Reproduce the graph on slide 3 of lesson 10 and add curves visualizing the estimated expected counts as a function of Biomass and with separate curves for the pH-levels. The curves should be based on the fitted model with interaction between Biomass and pH-level.

<div class="ans">

### Answers from students

<!-- #### Denis Tafintsev -->

```{r, eval = FALSE, include = FALSE}
with(species, {
  plot(Biomass, Species, type = "n")
  spp <- split(Species, pH)
  bio <- split(Biomass, pH)
  points(bio[[1]], spp[[1]], pch = 16, col = "red")
  points(bio[[2]], spp[[2]], pch = 16, col = "green")
  points(bio[[3]], spp[[3]], pch = 16, col = "blue")
  
  legend(x = "topright", legend = c("high", "medium", "low"), 
         pch = 16, col = c("red", "blue", "green"), title = "pH")
  
  curve(exp(3.76812 - 0.10713 * x), add = T, lwd = 2, col = "red")
  curve(exp(3.76812 - 0.33146 - (0.10713 + 0.03189) * x), 
        add = T, lwd = 2, col = "blue")
  curve(exp(3.76812 - 0.81557 - (0.10713 + 0.155028) * x), 
        add = T, lwd = 2, col = "green")
})
```

<!-- ### Camilo Esteban Silva Chiang -->

```{r, eval = FALSE, echo = FALSE}
# First we predict.
njlow <- 3.76 - 0.1 * species$Biomass - 0.81 - 0.155 * species$Biomass
njmid <- 3.76 - 0.1 * species$Biomass - 0.33 - 0.03 * species$Biomass
njhigh <- 3.76 - 0.1 * species$Biomass

# Now we do the plot
idx <- order(species$Biomass)
plot(Species ~ Biomass, data = species, col = pH, pch = 16, cex = 0.7)
points(exp(njlow)[idx] ~ species$Biomass[idx], type = "l", col = 2)
points(exp(njmid)[idx] ~ species$Biomass[idx], type = "l", col = 3)
points(exp(njhigh)[idx] ~ species$Biomass[idx], type = "l", col = 1)

# or optional
plot(Species ~ Biomass, data = species, col = pH, pch = 16, cex = 0.7)
for (i in 1:3) {
    test <- predict(mod3.glm, 
                    newdata = data.frame(Biomass = seq(0, 10, 0.1), 
                                         pH = levels(species$pH)[i]), 
                    type = "response")
    points(test ~ seq(0, 10, 0.1), type = "l", col = i)
}
```

</div>

<!--

## Taxonomy Data

We return to the taxonomy data used in the lecture on cluster analysis. In the file `taxonomy.Rdata` you find seven variables measured on plants from four different taxa. We will use multinomial regression as means for classification of taxa. A code file for performing cross-validation of classifications based on multinomial regression is found in the file `CV.class.multinom.R` on Fronter. Throughout this exercise you may need to increase the number of iterations used in the Maximum Likelihood estimation procedure. The default is 100 iterations. For example, in exercise a. `maxit = 200` should suffice. 

- a. Use multinomial regression to model the probabilities of the various taxa as a function of the variable `Sepal`.
Save the model as an object called `mod1.multi`. Use summary to display the results.

<div class="ans">

```{r}
mod1.multi <- multinom(Taxon ~ Sepal, data = taxonomy, maxit = 200)
summary(mod1.multi)
```

</div>

- b. What are the estimated linear predictors $\eta_j$ for each of the four taxa, $j=1,2,3,4$?

<div class="ans">

First we extract the estimated regression coefficients. Remember that for the reference level taxon the
regression coefficients are assumed equal to zero. We therefore add a row of zeros for this taxon (here `taxon = 1`).

```{r}
summod1.multi <- summary(mod1.multi)
co <- summod1.multi$coefficients
co <- rbind(c(0,0),co)
rownames(co)[1] <- "I"
co
```

This gives the following linear predictors:

$$\eta_1 = 0 + 0\cdot \text{Sepal} = 0$$

$$\eta_2 = 1.49 - 0.593\cdot \text{Sepal}$$

$$\eta_3 = 2.86 - 1.15\cdot \text{Sepal}$$

$$\eta_4 = -48.0 + 13.7\cdot \text{Sepal}$$

</div>

- c. What is the most probable taxon according to this model for a plant with average value of `Sepal`?
(Hint: If your coefficients are stored in a matrix called `co` you may compute the probabilities for each taxon by using the code `probs <- exp(co[,1]+co[,2]*x)/sum(exp(co[,1]+co[,2]*x))` where `x`is the value of
interest for `Sepal`.

<div class="ans">

```{r}
x <- mean(taxonomy$Sepal)
probs <- exp(co[ ,1] + co[ ,2] * x) / sum(exp(co[ ,1] + co[ ,2] * x))
probs
maxp <- which.max(probs)
maxp
```

We see that taxon I is the most probable with a probability equal to `r round(probs[maxp],3)`.

Extra: The following figure shows the probabilities of each taxon for various values of `Sepal`. Remember that these probabilities are the posterior probability estimates for the class in a classification terminology. We classify to the most probable class given the observed predictor values.

```{r}
xvector <- seq(0,7, by = 0.05)
probfunc <- function(x){
    exp(co[ ,1] + co[ ,2] * x) / sum(exp(co[ ,1] + co[ ,2] * x))
}
probmat <- t(sapply(xvector, probfunc))
matplot(xvector, probmat, lty = 1, type = "l", lwd = 2, 
        main = "Posterior probabilities for taxa", 
        xlab = "Sepal", ylab = "Estimated posterior probability")
legend(6, 0.8, legend = c("I", "II", "III", "IV"), lty = 1, col = 1:4, lwd = 2)
```

We observe that taxon III is the most probable for `Sepal<2.5`, taxon I for `2.5 < Sepal < 3.5` and taxon IV for `Sepal >= 3.5`. Taxon 2 is the most probable taxon only for `Sepal` $\approx$ `2.5`. 

</div>

- d. Execute the following plot command: `plot(Sepal ~ Taxon, data=taxonomy)`. Use the figure to explain the result from exercise c.

<div class="ans">

```{r}
plot(Sepal ~ Taxon, data = taxonomy)
```

We see that taxon I has the median (and probably also the mean) closest to the average Sepal length which is 3.0.
Taxa II and III are very similar, but slightly further away, hence their probabilities are a bit smaller. Taxa IV has
Sepal lengths much larger than the average and is thus the least probable taxon.

</div>

- e. Repeat the taxon classification of exercise c. by means of the `predict()` - function.

<div class="ans">

```{r}
predict(mod1.multi,newdata = data.frame(Sepal = mean(taxonomy$Sepal)))
```

</div>

- f. Use similar plots like in d. to identify other variables which are promising with regard to separating taxa I, II and III (Sepal is already a good variable to distinguish taxon IV from the others). Fit an extended model with the variables you have chosen and save the model as an object called `mod2.multi`.

<div class="ans">

I have found two variables which are promising; `Leaf` and `Petiole`, as shown in the figures below.

```{r}
plot(Leaf ~ Taxon, data = taxonomy)
plot(Petiole ~ Taxon, data = taxonomy)
```

`Leaf` seems to distinguish taxon III from the others, and `Petiole` may separate taxa I and II.

I therefore fit a model with `Sepal`, `Leaf` and `Petiole` as predictors for taxa. This model requires
many iterations in the estimation process. I set `maxit = 5000` to be sure. (`trace =FALSE` supresses
a long list of convergence details)

```{r}
mod2.multi <- multinom(Taxon ~ Sepal + Leaf + Petiole , data = taxonomy, 
                       maxit = 5000, trace = FALSE)
summary(mod2.multi)
```

</div>

-  Use a Chis-square test to test the joint significance of the extra variables you added 
as you extended `mod1.multi` to `mod2.multi`.

<div class="ans">

We test whether the difference in deviances between the two models is significant.

```{r}
dev.diff <- -2 * (logLik(mod1.multi) - logLik(mod2.multi))
pchisq(dev.diff, 6, lower = FALSE)
```

The p-value indicates that `Leaf` and `Petiole` are highly significant in addition to `Sepal` as
predictors in the model.

</div>

- h. Use the `predict()`function on `mod2.multi` with `newdata=taxonomy` to classify all plants in the
dataset to taxa. Then apply the `confusion()` function in the `mixlm`- package to compute the 
classification performance of the multinomial model. What is the accuracy and apparent classification error?

<div class="ans">

```{r, message = TRUE}
pred.all <- predict(mod2.multi, newdata = taxonomy)
mixlm::confusion(taxonomy$Taxon, pred.all)
```

I get an accuracy of 1 and an apparent classification error (APER) of 0. That is, a perfect "fit" to the data.
(Note the R-code `mixlm::` for extracting the confusion function from mixlm without loading the entire package).

</div>

- i. We should watch out for over-fitting. Use the `CV.class.multinom()` function to run a Leave-One-Out Cross-validation to validate the `mod2.multi` model.

<div class="ans">

```{r, message=FALSE}
#source("CV.class.multinom.R")
CVres <- CV.class.multinom(mod2.multi,data = taxonomy, trace = FALSE)
```

The cross-validation gave 5 mis-classifications. Taxon IV is perfectly classified, but there are some
mix-up's between taxa I, II and III. The accuracy is still high (0.96) and the total classification error is 
correspondingly small (0.04).

</div>

-->
