- Follow up the previous exercise by performing a Tukey test for all pair wise comparisons with overall 
(family-wise) error rate 5%. Give a summary of the results.

<div class="ans">

```{r}
NSRmod <- lm(Science ~ Age, data = NSRdata)
Anova(NSRmod, type = "III")
pt <- simple.glht(NSRmod, "Age", corr = c("Tukey"),level = 0.95)
print(pt)
```

The anova table indicates a clear significance of Age-group with regard to the interest level to Science subjects.
hence, at least two age-groups are have significantly different expected means. We can use the Tukey test to identify pairwise differences. 

The Tukey output provides differences in mean (column "center") between all pairs of groups, and at the top of output the minimum difference yielding a significant difference in means is given to be `r unique(round(with(pt$res, (Upper - Lower)/2), 4))`. That is, all age-groups with a difference in averages of more than `r unique(round(with(pt$res, (Upper - Lower)/2), 4))` are significantly different, according to Tukey. From the p-values we observe that `r sum(pt$res[,6] < 0.05)` out of `r dim(pt$res)[1]` pairs are significantly different, and the largest difference is found between age groups 30+ and 19-29.

```{r}
cld(pt)
```

The compact letter display gives a grouping of the similar levels, and there are three groups of levels that are internaly non-significantly different. 30+ is different from all other levels, whereas 1-12 and 16-19 are similar and 16-19, 13-15 and 19-29 are also similar.

</div>

## Barley Data

In `barley.rdata` are results from an experiment where the response is yield of barley pr 1000 square meter, and the factors sorts of barley, soil types, types of fertilizers. In addition was the experiment done in two different geographical areas (sites). 

- Assume a two factor model including the main effects of `sort` and `soil` and their interaction. State the model and explain all parameters under a sum-to-zero parametrization.

<div class="ans">

The model can be written as:

$$ y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \epsilon_{ijk}$$

where, $\epsilon \sim N(0, \sigma^2)$

The parametrization restrictions for "sum-to-zero"" are:

$$
\begin{aligned} 
\sum_i{\alpha_i} = 0, & \sum_j{\beta_j} = 0, & 
\sum_{i}{(\alpha\beta)_{ij}} = 0  \text{ and }
\sum_{j}{(\alpha\beta)_{ij}} = 0
\end{aligned}
$$

</div>

- Fit the model in R and estimate all parameters

<div class="ans">

```{r}
options(contrasts = c("contr.sum", "contr.poly"))
mod1 <- lm(Yield ~ sort * soil, data = Barley)
summod1 <- summary(mod1)
summod1$coef
summod1$sigma ^ 2
```

From the model summaries we find the estimated parameters to be:

1. The estimated overall mean yield: $\hat{\mu} = `r round(summod1$coefficients[1,1], 2)`$
2. The estimated effects of sort 1 and 2: $\hat{\alpha}_1=$ `r coef(mod1)[2]`, $\hat{\alpha}_2 = -\hat{\alpha}_1 =$ `r -coef(mod1)[2]`
3. The estimated effects of soil 1 and 2: $\hat{\beta}_1=$ `r coef(mod1)[3]`, $\hat{\beta}_2 = -\hat{\beta}_1 =$ `r -coef(mod1)[3]`
4. The estimated interaction effects of sort and soil: $\hat{\alpha\beta}_{11}=$ `r coef(mod1)[4]`, $\hat{\alpha\beta}_{12}=-\hat{\alpha\beta}_{11}$ `r -coef(mod1)[4]`,
$\hat{\alpha\beta}_{21}=-\hat{\alpha\beta}_{11}$ `r -coef(mod1)[4]`,
$\hat{\alpha\beta}_{22}=-\hat{\alpha\beta}_{21}=-\hat{\alpha\beta}_{12}=$ `r coef(mod1)[4]`
5. The estimated within `sort:soil` levels unexplained variability: $\hat{\sigma}^2=MSE=$ `r round(summod1$sigma^2, 2)`

</div>

- Make an interaction plot and try to conclude about the presence of interaction from the plot. How would you explain interaction effect in this example for a person with experience in agriculture, but minimal statistical experience? 

<div class="ans">

```{r, fig.asp=0.6, out.width='100%'}
plot(allEffects(mod1,confidence.level = .95))
```

From the plot, it seems that a farmer can expect an increase in yield when using sort 2 instead of sort 1 for both types of soil, and there may be a higher gain of sort 2 over sort 1 for soil-type 2 than for soil type 1. If this is so, there is a so-called interaction effect of sort and soil on yield. </div>

- Perform a hypothesis test for the interaction effect. Conclusion?

<div class="ans">

```{r}
Anova(mod1, type = "III")
```

The interaction between `soil` and `sort` is insignificant (high p-value). This result indicates that the slight non-parallel tendency of the lines in the interaction plot may be due to random errors. More data would be needed to verify any interaction effect.

</div>


- Fit a four factor model with sort, soil, fert and site as factors and with all possible 2-factor, 3-factor and 4-factor interactions. Use `Anova()` from the car-package with argument `type="III"` to produce an ANOVA table. Type III means that
all factors are tested as if they were the last factor to be added to the model.
Are there any higher order significant effects? Compare the R-squared and the adjusted R-squared values. What can you conclude from these?

<div class="ans">

```{r}
mod2 <- lm(Yield ~ sort * soil * fert * site, data = Barley)
Anova(mod2, type = "III")
summary(mod2)$r.squared
summary(mod2)$adj.r.squared
```

There is only one interaction effect significant at 5% levels, namely the 
`sort:fert:site` interaction. However, the least squares estimator is prone to inflated variance estimates and difficulties in finding significance when the number of variables `k` rises compared to the number of observations `N` and if variables are highy correlated. To many variables to estimate from a limited number of observations leaves few degrees of freedom to SSE, and tests with low power. Lack of significance may be an over-fitting problem. By removing some of the least significant variables, the problem is reduced, and a reduced model with significant effects may be identified. 

The large difference between $R^2$ and $R_{adj}^2$ is also an indication of an over-fitted model.

</div>

- Fit also a reduced model without site, but all other effects up to 3rd order
interactions. Perform a partial F-test with the `anova()` function to test whether site (and all its interactions with the others) should be excluded from the model (See also Exercise set 3 and Ex-3).

<div class="ans">

```{r}
mod3 <- lm(Yield ~ sort * soil * fert, data = Barley)
anova(mod2, mod3)
```

The `anova` result for partial F-test shows that site is needed to a certain degree in the model. The low p-value reject the hypothesis that there is no effect of `site` on `Yield`.

</div>
 
- In the `mixlm`-package there are convenient functions for performing automatic model selection by either `backward()` elimination of factors from a "full-model, by
`forward()` addition of factors from a minimal model with only intercept, or a combined `stepwise()` function which combines both forward and backward addition/elimination.

For the backward function the least significant factor is removed from the model in each step if the p-value is larger than testlevel `alpha`. If all factors are significant at any step, the procedure stops. The elimination obeys the so-called marginality principle (hierarchy of factors) which states that any lower order effect or interaction should not be removed from the model if it is part of a higher order significant interaction. This is a good principle for practical data analysis.

Try to run the `backward()` function on the model object you created above with all four factors and interactions. Use `alpha=0.05` as test level. Use the Step-information from the output to explain which factors being excluded at each step. What is the final reduced model?

<div class="ans">

```{r}
mod4 <- backward(mod2, alpha = 0.05)
```

In the first step, the 4th order interaction is removed. In 2nd, 3rd and 4th step some 3rd order and a 2nd order interactions are removed. After removing interaction between `sort`,`soil` and `site` in the 5th step, the resulting model is,

```{r, echo = FALSE}
mod4$call$formula
```

</div>
 
- Fit the reduced model and perform a model check using residual analysis.

<div class="ans">

The `backward` function has fitted the reduced model (her named `mod4`):

```{r, out.width='100%', fig.asp = 0.8}
mod4 <- lm(Yield ~ sort + soil + fert + site + sort:soil + 
     sort:fert + soil:fert + sort:site + fert:site + sort:fert:site, 
     data = Barley)

par(mfrow = c(2,2))
plot(mod4)
```

The residual plots give no evidence of any problems with the model assumptions.

</div>
 
- Make an interaction effects plot using the following code:

```{r effectPlot, eval = FALSE}
eff <- allEffects(mod4, confidence.level = .95)
plot(eff[3])
```

<div class="ans">

```{r effectPlot, eval = T, fig.asp = 0.7, out.width = '100%', echo = TRUE}
```

</div>

Try to explain the plot. Why do you think the interaction between sort, fertilizer and site was significant?

<div class="ans">

The plot shows that, when changing sort from 1 to 2, a farmer can expect different change in average Yield for `site:1` and `site:2` and the difference is not same when using `fertilizer:1` and `fertilizer:2`.

</div>
