# Analysis of variance

**We will use following datasets for this exercise:**

```{r, message=FALSE}
load("_data/city.rdata")
load("_data/nsr.rdata")
load("_data/barley.rdata")
```

**R-package we will use in this exercises**

```{r, message=FALSE, warning=FALSE}
# load the library
library(mixlm)
library(effects)
```

## Chlorine levels in cities

Independent measurements of chlorine (ppm parts per million) were taken from 3
large cities:

```{r}
city
```

- Load the city.rdata which is available on Fronter.

Before analyzing the data it is best to "stack" the data into two columns,  a response column `y` and a city factor column `city`. By using the `stack()` function in R you can restructure the city data by

```{r}
citydata <- stack(city)
```

- Use the `colnames()` function to rename the variables as `y` and `city`.

<div class="ans">

```{r}
colnames(citydata) <- c("y", "city")
```

</div>

- Assume an ANOVA-model with chlorine as response and city as a factor. What assumptions do you make?

<div class="ans">

The model assumptions are:

$$y_{ij} = \mu_i + \epsilon_{ij}$$

where the error terms are assumed independent and identically distributed $\epsilon_{ij} \sim N(0, \sigma^2)$. Further $\mu_i$ is the expected chlorine level in city $i$.

</div>

- Make a box-plot of the data. Describe the variabilities between cities and within cities.

<div class="ans">

```{r, out.width="60%"}
plot(y ~ city, data = citydata)
```

City 1 seems to have both the lowest average chlorine levels and the smallest variability in the measurements. Cities 2 and 3 are quite similar with both higher means and variability. 

</div>

- Fit the model and performa a residual analysis. Comment on the model fit.

<div class="ans">

```{r, out.width='100%', fig.asp = 0.6}
options(contrasts=c("contr.treatment", "contr.poly"))
citymod1 <- lm(y ~ city, data = citydata)
par(mfrow = c(1, 2))
plot(citymod1, which = c(1, 2))
```

The residual analysis reveals two problems: 1) Non-constant variability, which was also observed in the boxplot, and 2) A skeewed and non-normal distribution for the residuals with a heavy tail to the right.

</div>

- Make a log-transformation of $y$ to create the variable `logy`, and add the new variable to your data set by:

```{r}
citydata$logy <- log(citydata$y)
```

- Fit a new model with `logy` as response and check the model fit again.

<div class="ans">

```{r, out.width = '100%', fig.asp = 0.6}
citymod2 <- lm(logy~city, data = citydata)
par(mfrow = c(1,2))
plot(citymod2, which = c(1,2))
```

Both problems from the previous model appear to be corrected. We continue with this model.

</div>

- Estimate all model parameters from the latter model. Explain what the parameters measure given the paramtrization of your choice (See lecture notes from Feb 16th on parametrizations).

<div class="ans">

The ANOVA analysis using contr.treatment parametrization:

```{r}
summary(citymod2)
Anova(citymod2)
coefs <- citymod2$coefficients
```

We observe that $R^2 = 0.25$ indicating that there is still a lot of unexplained variability in the response. We get the following estimates for the city means of log(chlorine)-levels using this parametrization (city 1 is reference city):

$\mu_1 = \mu =$ `r round(coefs[1],2)`

$\mu_2 = \mu + \alpha_2 =$ `r round(coefs[1] + coefs[2],2)`

$\mu_3 = \mu + \alpha_3 =$ `r round(coefs[1] + coefs[3],2)`

The estimate of the error variance is $\hat{\sigma}^2 = MSE= SSE/(N-3) = 60.724/57 = 1.065$.

As we see from above, the "intercept" $\mu$  is the expected log-chlorine level for city 1. The $\alpha_2$ is the difference between city 1 and city 2, and $\alpha_3$ is the difference between city 1 and city 3. The $\sigma^2$ is the variance for log-chlorine levels measured in the same city.

</div>

* State the hypotheses for testing whether the expected chlorine levels differ between the cities, choose a test level $\alpha$ and perform the test. What is the conclusion?

<div class="ans">

The hypotheses are:

$$H_0: \mu_1 = \mu_2 = \mu_3$$ versus $$H_1:\mu_i \neq\mu_{i'}$$ for at least two different cities $i$ and $i'$.

Note:This is by the chosen parametrization equivalent to testing

$$H_0: \alpha_2 = \alpha_3 = 0$$ versus $$H_1:\alpha_i\neq0$$ for at least one $i \in \{2,3\}$.

For test-level $\alpha=0.05$ we reject the null-hypothesis if $F>F_{\alpha, 2, 57}$ or if the p-value is smaller than 0.05. Here we observe a very small p-value and reject the null-hypothesis. We claim that the expected log-chlorine levels differ between at least two of the cities, and from the observed means we know that cities 1 and 3 are significantly different since they have the largest observed difference in means. There may also be a significant difference between cities 1 and 2, and 2 and 3, but this should be checked by a pair-wise contrast.

</div>

## Data from the NSR education test

The Norwegian Centre for Science Recruitment (NSR) has an online "education test" where youths may answer a questionnaire to check their so-called cognitive types, their science interest, their preferred learning methods and their interest to various science subjects. The test suggests different ares within the STEM (Science, Technology, Engineering and Mathematics) within which the youth may find suitable work. 

We have an excerpt of these data which can be downloaded from Fronter as the `nsr.rdata` file. The data.frame `NSRdata` contains two variables, Science and Age:

```{r}
head(NSRdata,5)
```

`Science` is an average liking score (scale 1-6) to various STEM-subjects, and `Age` is a factor indicating different age-groups:

a) 1: 1-12 yrs
b) 13: 13-15 yrs
c) 16: 16-19 yrs
d) 19: 19-29 yrs
e) 30: 30 + yrs

- Perform an analysis of the NSR data to check whether Age influences the liking to STEM subjects. State the model, fit the model, check model assumptions, test hypotheses, and give model critique. Write a short summary of the results.


<div class="ans">

Answer will come later.

</div>
