# Group Exercises

## Exercise One ##

```{r, echo = F, message=FALSE}
library(mixlm)
library(car)
load("Audi.Rdata")
```

In the file Audi.Rdata in the Data folder on Fronter you find sales prices and technical data on 30 cars of type Audi A4. The data were collected on Feb 15th 2017. The variables are:

| Variable     |   | Description                                            |
|-------------:+:-:+:-------------------------------------------------------|
| `Price`      | : | Price of the car (In 1000 NOK)                         |
| `Km`         | : | Distance driven (in 1000 Km)                           |
| `Hk`         | : | Horse power                                            |
| `Transition` | : | Transition system (categorical). M=manual, A=Automatic |
| `Volume`     | : | Cylinder volume                                        |
| `Fuel`       | : | Fuel type (categorical). D=Diesel, G=Gasoline          |
| `CO2`        | : | CO2-emission (g/km)                                    |
| `Weight`     | : | The weight of the car                                  |
| `Year`       | : | Production year                                        |
| `Age`        | : | Years since production (=2017 – year)                  |

Consider Price as the response variable and all other variables (but not year) as candidate predictor variables.

a) Fit the full model including all candidate predictors. Write up the estimated model. Discuss the effect estimates. How should they be interpreted?

<div class = 'ans'>

The full model with all candidate predictors with price as response can be written as,

$$
\begin{aligned}
y = \beta_0 &+ \beta_1x_1 + \beta_2x_2 +  \beta_3x_3 + \beta_4x_4 + \beta_5x_5 \\ 
&+ \beta_6x_6 + \beta_7x_7 + \beta_8x_8 + \epsilon, \\
\text{ where, }\epsilon &\sim \text{NID}(0, \sigma^2)
\end{aligned}
$$ 

$$ $$
Here,

...

The least square estimate for this model can be obtained from R as,

```{r, echo = TRUE}
model.1 <- lm(Price ~ Km + Hk + Transition + Volume + Fuel + CO2 + Weight + Age, 
              data = Audi)
summary(model.1)
```

```{r, echo = FALSE}
cf <- round(coef(model.1), 2)
```

The estimated model can be written as,

$$
\begin{aligned}
\hat{y} = (`r cf[1]`) &+ (`r cf[2]`) x_1 + (`r cf[3]`) x_2 +  (`r cf[4]`) x_3 + (`r cf[5]`) x_4 \\
&+ (`r cf[6]`) x_5 + (`r cf[7]`) x_6 + (`r cf[8]`) x_7 + (`r cf[9]`) x_8
\end{aligned}
$$ 

</div>

b) Use variable selection methods or the best subsets approach to identify a good reduced model. Discuss the results.
c) Check the model assumptions of your final model.
d) Are there any influential observations?
e) Redo exercises a) to d), but start with a full model including `Age`, `Hk` and `Km` and all second order interactions between these variables. In R Commander you may specify the model like this:
    
    ```
    Price ~ (Age + Hk + Km)^2
    ```
    
    (_This notation means: Include all predictors up to second order interactions._) 
    Discuss the final model estimate.
f) If there is extra time: Can you find an even better model than the one you found in e?
